### 2020-12-23 Organismal_paper/Organismal_manuscript_and_code.Rmd
repo uptodate

Siphonophores are free-living predatory colonial hydrozoan cnidarians found in every ocean of the world. Siphonophore tentilla (tentacle side branches) are unique biological structures for prey capture, composed of a complex arrangement of cnidocytes (stinging cells) bearing different types of nematocysts (stinging capsules) and auxiliary structures. Tentilla present an extensive morphological and functional diversity across species. While associations between tentilla form and diet have been reported, the evolutionary history giving rise to this morphological diversity is largely unexplored. Here we examine the evolutionary gains and losses of novel tentillum substructures and nematocyst types on the most recent siphonophore phylogeny. Tentilla have a precisely coordinated high-speed strike mechanism of synchronous unwinding and nematocysts discharge. Here we characterize the kinematic diversity of this prey capture reaction using high-speed video and find relationships with morphological characters. Since tentillum discharge occurs in synchrony across a broad morphological diversity, we evaluate how phenotypic integration is maintaining character correlations across evolutionary time. We found that the tentillum morphospace has low dimensionality, we identified instances of heterochrony and morphological convergence, and generated hypotheses on the diets of understudied siphonophore species. Our findings indicate that siphonophore tentilla are phenotypically integrated structures with a complex evolutionary history leading to a phylogenetically structured diversity of forms which are predictive of kinematic performance and feeding habits.

@@

Siphonophores have fascinated zoologists for centuries for their extremely subspecialized colonial organization and integration. Today we have a comprehensive taxonomic coverage on the morphological diversity of this group due to the extensive work of siphonophore taxonomists in the past few decades  [@pugh1983benthic;@pugh1986new;@pugh1988two;@pugh2001review;@hissmann2005situ;@haddock2005re;@dunn2005marrus;@bardi2007taxonomic;@pugh2010three;@pugh2014review], which has been elegantly synthesized in detailed synopses [@totton1965synopsis, @mapstone2014global]. In addition, recent advances in phylogenetic analyses of siphonophores [@munro2018improved;@damian2020evolution] have provided a macroevolutionary context to interpret this diversity. With these assets in hand, we can now begin to study siphonophores from an orthogonal perspective, focusing on the diversity and evolutionary history of specific structures. Here we focus on one of such structures: the tentilla. Like many cnidarians, siphonophore tentacles bear side branches (tentilla) with nematocysts (Fig. \@ref(anatomy)). But unlike other cnidarians, most siphonophore tentilla are dynamic structures that react to prey encounters by rapidly unfolding the nematocyst battery to slap around the prey. This maximizes the surface area of contact between the nematocysts and the prey they fire upon. In addition, siphonophore tentilla present a remarkable diversity of morphologies (Fig. \@ref(figure2)), sizes, and nematocyst complements (Fig. \@ref(nematocysts)). Our overarching aim is to organize all this phenotypic diversity in a phylogenetic context, and identify the evolutionary processes that generated it.

![\label{anatomy} Siphonophore anatomy. A - Nanomia sp. siphonophore colony (photo by Catriona Munro). B, C - Illustration of a Nanomia colony, gastrozooid, and tentacle closeup (by Freya Goetz). D - Nanomia sp. Tentillum illustration and main parts. E - Differential interference contrast micrograph of the tentillum illustrated in D. Figure reproduced from Damian-Serrano et al. 2020 with permission. F. Action strip showing the behavior of tentilla during prey capture, illustrated by Riley Thompson. ](~/tentilla_morph/Organismal_paper/figures/anatomy.png)

In [@damian2020evolution], we collected the most extensive morphological dataset on siphonophore tentilla and nematocysts using state-of-the-art microscopy techniques, and expanded the taxon sampling of the phylogeny to disentangle the evolutionary history. The analyses we carried out led to novel, generalizable insights into the evolution of predatory specialization. The primary findings of that work were that generalists evolved from crustacean-specialist ancestors, and that feeding specializations were associated with distinct modes of evolution and character integration patterns. The work we present here is complementary to [@damian2020evolution], showcasing a far more detailed account on the evolutionary history of tentilla morphology.

@@

Nematocysts are unique biological weapons for defense and prey capture exclusive to the phylum Cnidaria. Mariscal ([@mariscal1974nematocysts]) reported that hydrozoans have the largest diversity of nematocyst types among cnidarians. Among them, siphonophores present the greatest variety of types [@mapstone2014global], and vary widely across taxa in which and how many types they carry on their tentacles (Fig. \@ref(nematocysts)). Werner [@werner1965nesselkapseln] noted that there are nine types of nematocyst found in siphonophores, of which four, anacrophore rhopalonemes, acrophore rhopalonemes, homotrichous anisorhizas, and birhopaloids, are unique to them. Heteroneme and haploneme nematocysts serve penetrant and entangling functions, while rhopalonemes and desmonemes work by adhering to the surface of the prey. While recent descriptive studies have expanded and confirmed our understanding of this diversity, the evolutionary history of nematocyst type gain and loss in siphonophores remains unexplored. Thus, here we reconstruct the evolution of shifts, gains, and losses of nematocyst types, subtypes, and other major categorical traits that led to the extant diversity we see in siphonophore tentilla.

@@

Distantly related organisms that evolved to feed on similar resources often evolve similar adaptations [@winemiller2015functional]. In [@damian2020evolution], we found strong associations between piscivory and haploneme shape across distantly related siphonophore lineages. These associations could have been produced by convergent changes in the adaptive optima of these characters. Here we set out to test this hypothesis using comparative model fitting methods. Analyzing the diversity of morphological states from a phylogenetic perspective allows us to identify the specific evolutionary processes that gave rise to it. Here we fit and compare a variety of macroevolutionary models to siphonophore tentilla morphology measurement data to identify instances of neutral divergence, stabilizing selection, changes in the speed of evolution, and convergent evolution.

In [@damian2020evolution] we fit discriminant analyses to identify characters that are predictive of feeding guild. These discriminant analyses can be used to generate hypotheses on the diets of ecologically understudied siphonophore species for which we have morphology data. Here we present a Bayesian prediction for the feeding guild of 45 species using the discriminant functions and morphological dataset in [@damian2020evolution]. As mentioned above, tentilla are far from being ornamental shapes and are in fact violently reactive weapons for prey capture [@Mackie:1987uy;@damian2020evolution]. While we now have detailed characterizations of tentilla morphologies across many species, the diversity of dynamic performances and their relationships to the undischarged morphologies have not been examined to date. To address this gap, we set out to record high-speed video of the *in vivo* discharge dynamics of several siphonophore species at sea, and compare the kinematic attributes to their morphological characters.

@@

All character data and the phylogeny analyzed here were published in [@damian2020evolution]. Details on the specimen collection, microscopy, and measurements can be found in the aforementioned publication. However, we included here the character definitions (SM15) and specimen list (SM16) in the Supporting Information. We log-transformed all the continuous characters that did not pass Shapiro-Wilks normality tests, and used the ultrametric constrained Bayesian time tree in all comparative analyses. Inapplicable characters were recorded as NA states, and species with states that could not be measured due to technical limitations were removed before the analyses. We used the feeding guild categories detailed in [@damian2020evolution] with one modification: including all *Forskalia* spp. as generalists instead of as a single *Forskalia* species on the tree after a reinterpretation of the data in [@purcell1981dietary]. In order to characterize the evolutionary history of tentilla morphology, we fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous in the R package *geiger* [@harmon2007geiger]. These models include a non-phylogenetic white-noise model (WN), a neutral divergence Brownian Motion model (BM), an early-burst decreasing rate model (EB), and an Ornstein-Uhlenbeck (OU) model with stabilizing selection around a fitted optimum trait value. In the same way as [@damian2020evolution] we then ordered the models by increasing parametric complexity (WN, BM, EB, OU), and compared their corrected Akaike Information Criterion (AICc) scores [@sugiura1978further]. We used the lowest (best) score using a delta cutoff of 2 units to determine significance relative to the next simplest model (SM10). We calculated model adequacy scores using the R package *arbutus* [@pennell2015model] (SM11), and calculated phylogenetic signals in each of the measured characters using Blomberg’s K [@blomberg2003testing] (SM10). To reconstruct the ancestral character states of nematocyst types and other categorical traits  (character matrix available in Supplementary Information), we used stochastic character mapping (SIMMAP) using the package *phytools* [@revell2012phytools].

In order to examine the phenotypic integration in the tentillum, we explored the correlational structure among continuous characters and among their evolutionary histories using principal component analysis (PCA) and phylogenetic PCA [@revell2012phytools]. Since the character dataset contains gaps due to missing data and inapplicable character states (SM14), we carried out these analyses on a subset of species and characters that allowed for the most complete dataset. This was done by removing the terminal filament characters (which are only shared by a small subset of species), and then removing species which had inapplicable states for the remaining characters (apolemiids and cystonects). In addition, we obtained the correlations between the phylogenetic independent contrasts [@felsenstein1985phylogenies] using the package *rphylip* [@revell2014rphylip] accounting for intraspecific variation. Using these contrasts, we identified multivariate correlational modules among characters. To test and quantify phenotypic integration between these multivariate modules, we used the phylogenetic phenotypic integration test in the package *geomorph* [@adams2016geomorph].

@@

In order to generate hypotheses on the diets of siphonophores using tentilla morphology, we used the discriminant analyses of principal components (DAPC) [@jombart2010discriminant] trained in [@damian2020evolution]. We predict the feeding guilds of species in the dataset for which there are no published feeding observations using their morphological data as inputs, and presenting the predictive output in the form of posterior probabilities for each guild category.

@@

*Evolutionary history of tentillum morphology* – In [@damian2020evolution], we produced the most speciose siphonophore molecular phylogeny to date, while incorporating the most recent findings in siphonophore deep node relationships. This phylogeny revealed for the first time that the genus *Erenna* is the sister to *Stephanomia amphytridis*. *Erenna* and *Stephanomia* bear the largest tentilla among all siphonophores, thus their monophyly indicates that there was a single evolutionary transition to giant tentilla. Siphonophore tentilla range in size from ~30 µm in some *Cordagalma* specimens to 2-4 cm in *Erenna* species, and up to 8 cm in *Stephanomia amphytridis* [@pugh2014review]. Most siphonophore tentilla measure between 175 and 1007 µm (1st and 3rd quartiles), with a median of 373 µm. The extreme gain of tentillum size in this newly found clade may have important implications for access to large prey size classes such as adult deep-sea fishes.

Siphonophore tentilla are defined as lateral, monostichous evaginations of the tentacle (including its gastrovascular lumen), armed with epidermal nematocysts [@totton1965synopsis]. The buttons on *Physalia* tentacles were not traditionally regarded as tentilla, but [@bardi2007taxonomic] and our observations [@munro2018improved], confirm that the buttons contain evaginations of the gastrovascular lumen, thus satisfying all the criteria for the definition. In this light, and given that most Cystonectae bear conspicuous tentilla, we conclude (in agreement with [@munro2018improved] and [@damian2020evolution]) that tentilla were present in the most recent common ancestor of all siphonophores, and secondarily lost twice, once in *Apolemia* and again in *Bathyphysa conifera*.In order to gain a broad perspective on the evolutionary history of tentilla, we reconstructed the phylogenetic positions of the main categorical character shifts using stochastic character mapping (SM1-9) and summarized in (Fig. \@ref(figure7)). Some of these characters include the gain and loss of nematocyst types.

@@

  Nematocyst type gain and loss is also associated with prey capture functions. For example, the loss of desmonemes and rhopalonemes in piscivorous *Erenna*, retaining solely the penetrant (and venom injecting) anisothizas and stenoteles (two size classes) is reminiscent of the two size classes of penetrant isorhizas in the fish-specialist cystonects. Moreover, with the gain of anisorhizas, desmonemes, and rhopalonemes, the Tendiculophora gained versatility in entangling and adhesive functions of the cnidoband and terminal filament, which may have allowed their feeding niches to diversify. Part of the effectiveness of calycophoran cnidobands at entangling crustaceans may be attributed to the subspecialization of their heteronemes. These shifted from the ancestral penetrating stenotele to the microbasic mastigophore (or eurytele in some species) with a long barmed shaft with many long spines. This heteroneme subtype could be better at interlocking with the setae of crustacean legs and antennae.

  In those species that have a functional terminal filament, the desmonemes and rhopalonemes play a fundamental role in the first stages of adhesion of the prey. In many species, the tugs of the struggling prey on the terminal filament trigger the cnidoband discharge [@Mackie:1987uy and pers. obs.].The adhesive terminal filament has been lost several times in the Euphysonectae (*Frillagalma*, *Lychnagalma*-*Physophora*, *Erenna*, and some species of *Cordagalma*). In these species, we hypothesize that a different trigger mechanism is at play, possibly involving the prey actively biting or grasping the tentillum or lure.

  The clades defined in [@damian2020evolution] are characterized by unique evolutionary innovations in their tentilla. The clade Eucladophora (containing Pyrostephidae, Euphysonectae, and Calycophorae) encompasses all of the extant Siphonophore species (178 of 186) except Cystonects and *Apolemia*. Innovations that arose along the stem of this group include spatially segregated heteroneme and haploneme nematocysts, terminal filaments, and elastic strands (Fig. \@ref(figure7)). Pyrostephids evolved a unique bifurcation of the axial gastrovascular canal of the tentillum known as the "saccus" [@totton1965synopsis]. The stem to the clade Tendiculophora (clade containing Euphysonectae and Calycophorae) subsequently acquired further novelties such as the desmoneme and rhopaloneme (acrophore subtype ancestral) nematocysts on the terminal filament (Fig. \@ref(figure7)), which bears no other nematocyst type. These are arranged in sets of 2 parallel rhopalonemes for each single desmoneme [@skaer1988formation;@skaer1991remodelling]. The involucrum is an expansion of the epidermal layer that can cover part or all of the cnidoband (Fig. \@ref(figure2)). This structure, together with differentiated larval tentilla, appeared in the stem branch to Clade A physonects. Calycophorans evolved novelties such as larger desmonemes at the distal end of the cnidoband, pleated pedicles with a “hood” (here considered homologous to the involucrum) at the proximal end of the tentillum, anacrophore rhopalonemes, and microbasic mastigophore-type heteronemes. While calycophorans have diversified into most of the extant described siphonophore species (108 of 186), their tentilla have not undergone any major categorical gains or losses since their most recent common ancestor. Nonetheless, they have evolved a wide variation in nematocyst and cnidoband sizes. Ancestrally (and retained in most prayomorphs and hippopodids), the calycophoran tentillum is recurved where the proximal and distal ends of the cnidoband are close together. Diphyomorph tentilla are slightly different in shape, with straighter cnidobands.

@@

*Evolution of tentillum and nematocyst characters* – Most (74%) characters present a significant phylogenetic signal, yet only total nematocyst volume, haploneme length, and heteroneme-to-cnidoband length ratio had a phylogenetic signal with K larger than 1 (SM10). Total nematocyst volume and cnidoband-to-heteroneme length ratio showed strongly conserved phylogenetic signals. The majority (67%) of characters were best fitted by BM models, indicating a history of neutral constant divergence. We did not find any relationship between phylogenetic signal and specific model support, where characters with high and low phylogenetic signal were broadly distributed among the best fitted for each model. One-third of the characters measured in [@damian2020evolution] did not recover significant support for any of the phylogenetic models tested, indicating they are either not phylogenetically conserved, or they evolved under a complex evolutionary process not represented among the models tested (SM10). Haploneme nematocyst length was the only character with support for an EB model of decreasing rate of evolution with time. No character had support for a single-optimum OU model (when not informed by feeding guild regime priors). The model adequacy tests (SM11) indicate that many characters may have a relationship between the states and the rates of evolution (Sasr) not captured in the basic models compared here, accompanied by a signal of unaccounted rate heterogeneity (Cvar). No characters show significant deviations in the overall rate of evolution estimated (Msig). Some characters show a perfect fit (no significant deviations across all metrics) under BM evolution, such as heteroneme shape, length, width & volume, haploneme width & SA/V, tentacle width and pedicle width. Haploneme row number and rhopaloneme shape have significant deviations across four metrics, indicating that BM (best model) is a poor fit. These characters likely evolved under complex models which would require many more data points than we have available to fit with accuracy.

@@

![\label{surface} SURFACE plots showing convergent evolutionary regimes modelled under OU for (A) haploneme nematocyst length & width, and (B) for PC1 & 2 of all continuous characters with the exception of terminal filament nematocysts, and removing taxa with inapplicable character states. Node numbers on the tree label different regimes, regimes of the same color are identified as convergent. Small circles on the scatterplots indicate species values, large circles indicate the average position of the OU optima (*theta*) for a given combination of convergent regimes.](~/tentilla_morph/Organismal_paper/figures/SURFACE-01.jpg){height=700px}

>>>>

Siphonophores are free-living predatory colonial hydrozoan cnidarians found in every ocean of the world. Siphonophore tentilla (tentacle side branches) are unique biological structures for prey capture, composed of a complex arrangement of cnidocytes (stinging cells) bearing different types of nematocysts (stinging capsules) and auxiliary structures. Tentilla present an extensive morphological and functional diversity across species. While associations between tentilla form and diet have been reported, the evolutionary history giving rise to this morphological diversity is largely unexplored. Here we examine the evolutionary gains and losses of novel tentillum substructures and nematocyst types on the most recent siphonophore phylogeny. Tentilla have a precisely coordinated high-speed strike mechanism of synchronous unwinding and nematocysts discharge. Here we characterize the kinematic diversity of this prey capture reaction using high-speed video and find relationships with morphological characters. Since tentillum discharge occurs in synchrony across a broad morphological diversity, we evaluate how phenotypic integration is maintaining character correlations across evolutionary time. We found that the tentillum morphospace has low dimensionality, we identified instances of heterochrony and morphological convergence, and generated hypotheses on the diets of understudied siphonophore species. Our findings indicate that siphonophore tentilla are phenotypically integrated structures with a complex evolutionary history leading to a phylogenetically-structured diversity of forms which are predictive of kinematic performance and feeding habits.

@@

Siphonophores have fascinated zoologists for centuries for their extremely subspecialized colonial organization and integration. Today we have a comprehensive taxonomic coverage on the morphological diversity of this group due to the extensive work of siphonophore taxonomists in the past few decades  [@pugh1983benthic;@pugh1986new;@pugh1988two;@pugh2001review;@hissmann2005situ;@haddock2005re;@dunn2005marrus;@bardi2007taxonomic;@pugh2010three;@pugh2014review], which has been elegantly synthesized in detailed synopses [@totton1965synopsis;@mapstone2014global]. In addition, recent advances in phylogenetic analyses of siphonophores [@munro2018improved;@damian2020evolution] have provided a macroevolutionary context to interpret this diversity. With these assets in hand, we can now begin to study siphonophores from an orthogonal perspective, focusing on the diversity and evolutionary history of specific structures. Here we focus on one of such structures: the tentilla. Like many cnidarians, siphonophore tentacles bear side branches (tentilla) with nematocysts (Fig. \@ref(anatomy)C-E). But unlike other cnidarians, most siphonophore tentilla are dynamic structures that react to prey encounters by rapidly unfolding the nematocyst battery to slap around the prey (Fig. \@ref(anatomy)F). This maximizes the surface area of contact between the nematocysts and the prey they fire upon. In addition, siphonophore tentilla present a remarkable diversity of morphologies (Fig. \@ref(figure2)), sizes, and nematocyst complements (Fig. \@ref(nematocysts)). Our overarching aim is to organize all this phenotypic diversity in a phylogenetic context, and identify the evolutionary processes that generated it.

![\label{anatomy} Siphonophore anatomy. A - Nanomia sp. siphonophore colony (photo by Catriona Munro). B, C - Illustration of a Nanomia colony, gastrozooid, and tentacle closeup (by Freya Goetz). D - Nanomia sp. Tentillum illustration and main parts. E - Differential interference contrast micrograph of the tentillum illustrated in D. Figure reproduced from Damian-Serrano et al. 2020 with permission. F. Action strip showing the behavior of tentilla during prey capture, illustrated by Riley Thompson. ](~/tentilla_morph/Organismal_paper/figures/anatomy-01.png)

In @damian2020evolution, we collected the most extensive morphological dataset on siphonophore tentilla and nematocysts using state-of-the-art microscopy techniques, and expanded the taxon sampling of the phylogeny to disentangle the evolutionary history. The analyses we carried out led to novel, generalizable insights into the evolution of predatory specialization. The primary findings of that work were that generalists evolved from crustacean-specialist ancestors, and that feeding specializations were associated with distinct modes of evolution and character integration patterns. The work we present here is complementary to @damian2020evolution, showcasing a far more detailed account on the evolutionary history of tentilla morphology.

@@

Nematocysts are unique biological weapons for defense and prey capture exclusive to the phylum Cnidaria. @mariscal1974nematocysts reported that hydrozoans have the largest diversity of nematocyst types among cnidarians. Among them, siphonophores present the greatest variety of types [@mapstone2014global], and vary widely across taxa in which and how many types they carry on their tentacles (Fig. \@ref(nematocysts)). @werner1965nesselkapseln noted that there are nine types of nematocyst found in siphonophores, of which four, anacrophore rhopalonemes, acrophore rhopalonemes, homotrichous anisorhizas, and birhopaloids, are unique to them. Heteroneme and haploneme nematocysts serve penetrant and entangling functions, while rhopalonemes and desmonemes work by adhering to the surface of the prey. While recent descriptive studies have expanded and confirmed our understanding of this diversity, the evolutionary history of nematocyst type gain and loss in siphonophores remains unexplored. Thus, here we reconstruct the evolution of shifts, gains, and losses of nematocyst types, subtypes, and other major categorical traits that led to the extant diversity we see in siphonophore tentilla.

@@

Distantly related organisms that evolved to feed on similar resources often evolve similar adaptations [@winemiller2015functional]. In @damian2020evolution, we found strong associations between piscivory and haploneme shape across distantly related siphonophore lineages. These associations could have been produced by convergent changes in the adaptive optima of these characters. Here we set out to test this hypothesis using comparative model fitting methods. Analyzing the diversity of morphological states from a phylogenetic perspective allows us to identify the specific evolutionary processes that gave rise to it. Here we fit and compare a variety of macroevolutionary models to siphonophore tentilla morphology measurement data to identify instances of neutral divergence, stabilizing selection, changes in the speed of evolution, and convergent evolution.

In @damian2020evolution we fit discriminant analyses to identify characters that are predictive of feeding guild. These discriminant analyses can be used to generate hypotheses on the diets of ecologically understudied siphonophore species for which we have morphology data. Here we present a Bayesian prediction for the feeding guild of 45 species using the discriminant functions and morphological dataset in @damian2020evolution. As mentioned above, tentilla are far from being ornamental shapes and are in fact violently reactive weapons for prey capture [@Mackie:1987uy;@damian2020evolution]. While we now have detailed characterizations of tentilla morphologies across many species, the diversity of dynamic performances and their relationships to the undischarged morphologies have not been examined to date. To address this gap, we set out to record high-speed video of the *in vivo* discharge dynamics of several siphonophore species at sea, and compare the kinematic attributes to their morphological characters.

@@

All character data and the phylogeny analyzed here were published in @damian2020evolution. Details on the specimen collection, microscopy, and measurements can be found in the aforementioned publication. To facilitate access, we re-included here the character definitions (SM15) and specimen list (SM16) in the Supporting Information. We log-transformed all the continuous characters that did not pass Shapiro-Wilks normality tests, and used the ultrametric constrained Bayesian time tree in all comparative analyses. Inapplicable characters were recorded as NA states, and species with states that could not be measured due to technical limitations were removed before the analyses. We used the feeding guild categories detailed in @damian2020evolution with one modification: including all *Forskalia* spp. as generalists instead of as a single *Forskalia* species on the tree after a reinterpretation of the data in @purcell1981dietary. In order to characterize the evolutionary history of tentilla morphology, we fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous in the R package *geiger* [@harmon2007geiger]. These models include a non-phylogenetic white-noise model (WN), a neutral divergence Brownian Motion model (BM), an early-burst decreasing rate model (EB), and an Ornstein-Uhlenbeck (OU) model with stabilizing selection around a fitted optimum trait value. In the same way as @damian2020evolution, we then ordered the models by increasing parametric complexity (WN, BM, EB, OU), and compared their corrected Akaike Information Criterion (AICc) scores [@sugiura1978further]. We used the lowest (best) score using a delta cutoff of 2 units to determine significance relative to the next simplest model (SM10). We calculated model adequacy scores using the R package *arbutus* [@pennell2015model] (SM11), and calculated phylogenetic signals in each of the measured characters using Blomberg’s K [@blomberg2003testing] (SM10). To reconstruct the ancestral character states of nematocyst types and other categorical traits  (character matrix available in Supplementary Information), we used stochastic character mapping (SIMMAP) using the package *phytools* [@revell2012phytools].

In order to examine the degree of phenotypic integration within the tentillum, we explored the correlational structure among continuous characters and among their evolutionary histories using principal component analysis (PCA) and phylogenetic PCA [@revell2012phytools]. Since the character dataset contains gaps due to missing data and inapplicable character states (SM14), we carried out these analyses on a subset of species and characters that allowed for the most complete dataset. This was done by removing the terminal filament characters (which are only shared by a small subset of species), and then removing species which had inapplicable states for the remaining characters (apolemiids and cystonects). In addition, we obtained the correlations between the phylogenetic independent contrasts [@felsenstein1985phylogenies] using the package *rphylip* [@revell2014rphylip] accounting for intraspecific variation. Using these contrasts, we identified multivariate correlational modules among characters. To test and quantify phenotypic integration between these multivariate modules, we used the phylogenetic phenotypic integration test in the package *geomorph* [@adams2016geomorph].

@@

In order to generate hypotheses on the diets of siphonophores using tentilla morphology, we used the discriminant analyses of principal components (DAPC) [@jombart2010discriminant] trained in @damian2020evolution. We predict the feeding guilds of species in the dataset for which there are no published feeding observations using their morphological data as inputs, and presenting the predictive output in the form of posterior probabilities for each guild category.

@@

*Evolutionary history of tentillum morphology* – In @damian2020evolution, we produced the most speciose siphonophore molecular phylogeny to date, while incorporating the most recent findings in siphonophore deep node relationships. This phylogeny revealed for the first time that the genus *Erenna* is the sister to *Stephanomia amphytridis*. *Erenna* and *Stephanomia* bear the largest tentilla among all siphonophores, thus their monophyly indicates that there was a single evolutionary transition to giant tentilla. Siphonophore tentilla range in size from ~30 µm in some *Cordagalma* specimens to 2-4 cm in *Erenna* species, and up to 8 cm in *Stephanomia amphytridis* [@pugh2014review]. Most siphonophore tentilla measure between 175 and 1007 µm (1st and 3rd quartiles), with a median of 373 µm. The extreme gain of tentillum size in this newly found clade may have important implications for access to large prey size classes such as adult deep-sea fishes.

Siphonophore tentilla are defined as lateral, monostichous evaginations of the tentacle (including its gastrovascular lumen), armed with epidermal nematocysts [@totton1965synopsis]. The buttons on *Physalia* tentacles were not traditionally regarded as tentilla, but @bardi2007taxonomic, @munro2018improved, and our own observations confirm that the buttons contain evaginations of the gastrovascular lumen, thus satisfying all the criteria for the definition. In this light, and given that most Cystonectae bear conspicuous tentilla, we conclude (in agreement with @munro2018improved and @damian2020evolution) that tentilla were present in the most recent common ancestor of all siphonophores, and secondarily lost twice, once in *Apolemia* and again in *Bathyphysa conifera*. In order to gain a broad perspective on the evolutionary history of tentilla, we reconstructed the phylogenetic positions of the main categorical character shifts (such as gains and losses of nematocyst types) using stochastic character mapping (SM1-9) and manual reconstructions. This phylogenetic roadmap of evolutionary novelties is summarized in (Fig. \@ref(figure7)).

@@

  Nematocyst type gain and loss is also associated with prey capture functions. For example, the loss of desmonemes and rhopalonemes in piscivorous *Erenna*, retaining solely the penetrant (and venom injecting) anisothizas and stenoteles (two size classes) is reminiscent of the two size classes of penetrant isorhizas in the fish-specialist cystonects. Moreover, with the gain of anisorhizas, desmonemes, and rhopalonemes, the Tendiculophora gained versatility in entangling and adhesive functions of the cnidoband and terminal filament, which may have allowed their feeding niches to diversify. Part of the effectiveness of calycophoran cnidobands at entangling crustaceans may be attributed to the subspecialization of their heteronemes. These shifted from the ancestral penetrating stenotele to the microbasic mastigophore (or eurytele in some species) with a long barbed shaft armed with many long spines. This heteroneme subtype could be better at interlocking with the setae of crustacean legs and antennae.

  In those species that have a functional terminal filament, the desmonemes and rhopalonemes play a fundamental role in the first stages of adhesion of the prey. In many species, the tugs of the struggling prey on the terminal filament trigger the cnidoband discharge [@Mackie:1987uy and pers. obs.]. The adhesive terminal filament has been lost several times in the Euphysonectae (*Frillagalma*, *Lychnagalma*-*Physophora*, *Erenna*, and some species of *Cordagalma*). In these species, we hypothesize that a different trigger mechanism is at play, possibly involving the prey actively biting or grasping the tentillum or lure.

  The clades defined in @damian2020evolution are characterized by unique evolutionary innovations in their tentilla. The clade Eucladophora (containing Pyrostephidae, Euphysonectae, and Calycophorae) encompasses all of the extant Siphonophore species (178 of 186) except Cystonects and *Apolemia*. Innovations that arose along the stem of this group include spatially segregated heteroneme and haploneme nematocysts, terminal filaments, and elastic strands (Fig. \@ref(figure7)). Pyrostephids evolved a unique bifurcation of the axial gastrovascular canal of the tentillum known as the "saccus" [@totton1965synopsis]. The stem to the clade Tendiculophora (clade containing Euphysonectae and Calycophorae) subsequently acquired further novelties such as the desmoneme and rhopaloneme (acrophore subtype ancestral) nematocysts on the terminal filament (Fig. \@ref(figure7)), which bears no other nematocyst type. These are arranged in sets of 2 parallel rhopalonemes for each single desmoneme [@skaer1988formation;@skaer1991remodelling]. The involucrum is an expansion of the epidermal layer that can cover part or all of the cnidoband (Fig. \@ref(figure2)). This structure, together with differentiated larval tentilla, appeared in the stem branch to Clade A physonects. Calycophorans evolved novelties such as larger desmonemes at the distal end of the cnidoband, pleated pedicles with a “hood” (here considered homologous to the involucrum) at the proximal end of the tentillum, anacrophore rhopalonemes, and microbasic mastigophore-type heteronemes. While calycophorans have diversified into most of the extant described siphonophore species (108 of 186), their tentilla have not undergone any major categorical gains or losses since their most recent common ancestor. Nonetheless, they have evolved a wide variation in nematocyst and cnidoband sizes. Ancestrally (and retained in most prayomorphs and hippopodids), the calycophoran tentillum is recurved where the proximal and distal ends of the cnidoband are close together. Diphyomorph tentilla are slightly different in shape, with straighter cnidobands.

@@

*Evolution of tentillum and nematocyst characters* – Most (74%) characters present a significant phylogenetic signal, yet only total nematocyst volume, haploneme length, and heteroneme-to-cnidoband length ratio had a phylogenetic signal with K larger than 1 (SM10). Total nematocyst volume and cnidoband-to-heteroneme length ratio showed strongly conserved phylogenetic signals. The majority (67%) of characters were best fitted by BM models, indicating a history of neutral constant divergence. We did not find any relationship between phylogenetic signal and specific model support, where characters with high and low phylogenetic signal were broadly distributed among the best fitted for each model. One-third of the characters measured in @damian2020evolution did not recover significant support for any of the phylogenetic models tested, indicating they are either not phylogenetically conserved, or they evolved under a complex evolutionary process not represented among the models tested (SM10). Haploneme nematocyst length was the only character with support for an EB model of decreasing rate of evolution with time. No character had support for a single-optimum OU model (when not informed by feeding guild regime priors). The model adequacy tests (SM11) indicate that many characters may have a relationship between the states and the rates of evolution (Sasr) not captured in the basic models compared here, accompanied by a signal of unaccounted rate heterogeneity (Cvar). No characters show significant deviations in the overall rate of evolution estimated (Msig). Some characters show a perfect fit (no significant deviations across all metrics) under BM evolution, such as heteroneme shape, length, width & volume, haploneme width & SA/V, tentacle width and pedicle width. Haploneme row number and rhopaloneme shape have significant deviations across four metrics, indicating that BM (best model) is a poor fit. These characters likely evolved under complex models which would require many more data points than we have available to fit with accuracy.

*Phenotypic integration of the tentillum* – Phenotypically integrated structures maintain evolutionary correlations between its constituent characters. Of the phylogenetic correlations among tentillum and nematocyst characters examined here (Fig. \@ref(figure8)a, lower triangle), 81.3% were positive and 18.7% were negative, while of the ordinary correlations (Fig. \@ref(figure8)a, upper triangle) 74.6% were positive and 25.4% were negative. Half (49.9%) of phylogenetic correlations were >0.5, while only 3.6% are < -0.5. Similarly, among the correlations across extant species, 49.1% were >0.5 and only 1.5% were < -0.5. In addition, we found that 13.9% of character pairs had opposing phylogenetic and ordinary correlation coefficients (Fig. \@ref(figure8)B). Just 4% of character pairs have negative phylogenetic and positive ordinary correlations (such as rhopaloneme elongation ~ heteroneme-to-cnidoband length ratio and haploneme elongation, or haploneme elongation ~ heteroneme number), and only 9.9% of character pairs had positive phylogenetic correlation yet negative ordinary correlation (such as heteroneme elongation ~ cnidoband convolution and involucrum length, or rhopaloneme elongation with cnidoband length). These disparities could be explained by Simpson’s paradox [@blyth1972simpson]: the reversal of the sign of a relationship when a third variable (or a phylogenetic topology, as suggested by @uyeda2018rethinking) is considered. However, no character pair had correlation coefficient differences larger than 0.64 between ordinary and phylogenetic correlations (heteroneme shaft extension ~ rhopaloneme elongation has a Pearson’s correlation of 0.10 and a phylogenetic correlation of -0.54). Rhopaloneme elongation shows the most incongruencies between phylogenetic and ordinary correlations with other characters. We identified four hypothetical modules among the tentillum characters: (1) The tentillum scaffold module including cnidoband length & width, nematocyst row number, pedicle & elastic strand width, tentacle width; (2) the heteroneme module including heteroneme length & width, shafts length & width; (3) the haploneme module including length and width; and (4) the terminal filament module including desmoneme & rhopaloneme length & width. The phenotypic integration test showed significant integration signal between all modules, tentillum and haploneme modules sharing the greatest regression coefficient (SM12).

@@

*Convergent evolution* – Convergence is a widespread evolutionary phenomenon where distantly related clades independently evolve similar phenotypes. When the dimensionality of the state space is small as it is in tentilla morphology, convergence is more likely given the same amount of evolutionary change. Using the package SURFACE [@ingram2013surface], we identified convergence in haploneme nematocyst shape and in morphospace position. In @damian2020evolution, we identified haploneme nematocyst shape as one of the traits associated with the convergent evolution of piscivory. Here we find that indeed wider haploneme nematocysts have convergently evolved in the piscivore cytonects and *Erenna* spp. (Fig. \@ref(surface)A). Extremely narrow haplonemes have also evolved convergently in *Cordagalma ordinatum* and copepod specialist calycophorans such as *Sphaeronectes koellikeri*. When integrating many traits into a couple principal components, we find two distinct convergences between euphysonects and calycophorans with a reduced prey capture apparatus. Those convergences are between *Frillagalma vityazi* and calycophorans, and once again between *Cordagalma ordinatum* and *Spaheronectes koellikeri* (Fig. \@ref(surface)B).
